---
title: "Predicting risk of Coronary Heart Disease using Machine Learning"
author: "Dennis Wiersma"
date: "2022-10-05"
urlcolor: blue
output: 
  pdf_document: 
    highlight: kate
    number_sections: yes
    citation_package: natbib
    
    includes: 
      in_header: preamble.tex
      before_body: prefix.tex
      
bibliography: references.bib
---

```{r, include=FALSE}
# Copyright (c) 2022 Dennis Wiersma.
# Licensed under GPLv3. See LICENSE file.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Excludes all code from the report
knitr::opts_chunk$set(include = FALSE)
```

```{r loadingDataset}
# Load dataset
dataset <- read.table(file = "../data/processedData.csv", header = TRUE, sep = ",", stringsAsFactors = TRUE)
```

\newpage


# Introduction
Heart disease is a condition that gets more prevalent by the day. It is one of the most, if not the most, occurring cause of death in this day and age [@World_Health_Organization_2020_who]. It is therefore important to know what factors can lead to someone developing heart disease. Thankfully, a lot is already known about what goes into developing this condition. So, if it is known what causes heart disease, then what can we do with this information? The answer is prevention. Possibly just as important as knowing what causes heart disease would be to know whether someone is likely to develop a heart condition in the near future. After all, if one knows they're at risk of succumbing to the number one killer in the world, they can still make an effort to subvert disaster. In the process of gathering such information a patient could seek advice from their physician, who could then evaluate their lifestyle and physical condition. Using this information the medical practitioner can then estimate whether the patient is likely to be subject to the unease of heart disease. \
Computer scientists wouldn't be computer scientists if they weren't trying to program someone out of a job, so what if this evaluation by a doctor could be automated by a computer. A piece of software could evaluate data from dozens of patients in mere seconds. This would cause the need for a physician to greatly decrease for this procedure, lowering both cost and barrier to entry. \
To achieve this dream of automation a range of techniques were used, the most important of which is ML (Machine Learning) which is a sub field of AI (Artificial Intelligence). Using these ML techniques a model can be trained using a large amount of data. This model can then predict certain values for a given input. For example, such a model could predict whether a patient has the risk of developing heart disease when given information about that patient's medical condition. \
As mentioned before, for the training of this ML model a rather large amount of data is required. The dataset which was used for completing this training phase was gathered from the FHS (Framingham Heart Study) and contains information on the risk of subjects developing CHD (Coronary Heart Disease) within the next ten years. Since these models can be fairly specific to the data on which they have been trained, any bias in the dataset will also translate into the model produced from it. This therefore leaves us with the following question to answer: How well can one predict 10 year risk of coronary heart disease in subjects from the Framingham Heart Study?

\newpage

# Methods
\newpage

# Results
## Data acquisition & preprocessing

## Validation & Performance

## Research findings

\newpage

# Conclusion & Discussion

## Result based conclusions

## Discussion

## General conclusions and perspective

\newpage

# Bibliography

[@Michiel_Noback_2022_hanze]
